---
title: "QPot: An R Package for Stochastic Differential Equation Quasi-Potential Analysis"
author:
  - "Christopher Moore, Chris Stieha, Ben Nolting, Maria Cameron, and Karen Abbott"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{1. Analyzing the deterministic skeleton}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Analyzing the deterministic skeleton
========================================================
## Example 1 from Moore et al. (2016)
### A system of equations
In Moore et al. (2016) we use an example of a coupled consumer-resource equation developed for plankton, $x$, and their consumers, $y$:

$$\frac{\mathrm{d}x(t)}{\mathrm{d}t} = \alpha x(t)\left(1 - \frac{x(t)}{\beta}\right) - \frac{\delta x^2(t)y(t)}{\kappa + x^2(t)}$$
$$\frac{\mathrm{d}y(t)}{\mathrm{d}t} = \frac{\gamma x^2(t)y(t)}{\kappa + x^2(t)} - \mu y^2(t)$$

For this system, we use a specific parametrization to generate two non-trivial stable equilibria.  Specifically, we use the parameters:

Parameter  | Value    | Biological description
---------- | :------- | :----------------------------------------
$\alpha$   | 1.54     | Maximal growth rate of plankton
$\beta$    | 10.14    | Carrying capacity of plankton
$\delta$   | 1.0      | Maximal feeding rate of the consumers
$\gamma$   | 0.476    | Conversion rate of plankton to consumer
$\kappa$   | 1.0      | Half-consumption saturation rate
$\mu$      | 0.112509 | Death rate of the consumer

### Visualizing dynamics
We can create a vector field of the deterministic skeleton by using package [phaseR](https://cran.r-project.org/web/packages/phaseR/index.html).  First, we will load [phaseR](https://cran.r-project.org/web/packages/phaseR/index.html) and its dependency, [deSolve](https://cran.r-project.org/web/packages/deSolve/index.html) ([phaseR](https://cran.r-project.org/web/packages/phaseR/index.html) makes use of `deSolve::ode`):

```{r, message = F}
library(package = "deSolve")
library(package = "phaseR")
```

Second, we write our equations above in the pseudo-code format:

```
	model <- function(time, initial conditions, parameters){
	assign state variables to initial conditions
	assign parameters
	create an object to store output
	equations
	a list as the output
	}
```

in R as

```{r}
model.ex1 <- function(t, y, parameters) {
  x <- y[1]
  y <- y[2]
  alpha <- parameters["alpha"]
  beta <- parameters["beta"]
  delta <- parameters["delta"]
  kappa <- parameters["kappa"]
  gamma <- parameters["gamma"]
  mu <- parameters["mu"]
  dy <- numeric(2)
  dy[1] <- (alpha * x) * (1 - (x / beta)) - ((delta * (x^2) * y) / (kappa + (x^2)))
  dy[2] <- ((gamma * (x^2) * y) / (kappa + (x^2))) - mu * (y^2)
  list(dy)
}
```
Then, we plot the direction field and the zero-growth isoclines (i.e., nullclines):

```{r}
model.parms <- c("alpha" = 1.54, "beta" = 10.14, "delta" = 1, "gamma" = 0.476, "mu" = 0.112509, "kappa" = 1)
xlims <- c(0, 6)
ylims <- c(0, 6)
flowField(deriv = model.ex1, xlim = xlims, ylim = ylims, parameters = model.parms, points = 30, add = FALSE, state.names = c("x", "y")) |> plot()
nullclines(model.ex1, xlim = xlims, ylim = ylims, parameters = model.parms, points = 250, col = c("blue", "red"), state.names = c("x", "y")) |> plot()
```

### Identifying and classifying equilibria
We can see from the field that trajectories may take many paths to different areas or points in phase space (i.e., there are multiple basins of attraction).  But more reliably, we can see that the nullclines cross several times, which means that at those points we have equilibria.  For the purposes of our example, we are interested in interior points (i.e., $x$ and $y$ have populations > 0).

There are several ways to find solutions and classify equilibrium points.  Users are encouraged to familiarize themselves with [CRAN Task View: Differential Equations](https://cran.r-project.org/web/views/DifferentialEquations.html) for a summary of available packages.  Here, we use package [rootSolve](https://cran.r-project.org/web/packages/rootSolve/index.html) to find equilibria.  At each equilibrium point, we also classify the behavior about the point by finding the eigenvalues of the Jacobian matrix, known as linear or local stability analysis.

#### Finding equilibria
For simple models, equilibria can be found analytically.  But for most non-linear models, a solver must be used.  Because we have an idea where the equilibria are, we create an area to find steady-state equilibria using `rootSolve::stode`.  Because it's a small space, we'll first use a `for` loop over the x- and y-area we wish to sample, then we find unique values of (x, y) to give us a matrix of unique equilibria.
```{r}
library(package = "rootSolve")
xspace <- seq(from = 1, to = 5, length.out = 10)
yspace <- seq(from = 2.5, to = 4, length.out = 10)
l.xspace <- length(x = xspace)
l.yspace <- length(x = yspace)
space.mat <- matrix(data = NA, nrow = l.xspace * l.yspace, ncol = 2)

for (i in 1:l.xspace) {
  for (j in 1:l.yspace) {
    y <- c(x = xspace[i], y = yspace[j])
    STO <- stode(y = y, func = model.ex1, parms = model.parms, positive = T)
    space.mat[(((i - 1) * l.xspace) + j), ] <- STO$y
  }
}
eqs <- unique(x = round(x = space.mat, digits = 3))
```
And, for further confirmation, if the plot window is still open, we can add them to ensure they look correct:
```{r}
plot(x = eqs[, 1], y = eqs[, 2], cex = 1.5)
```

#### Classifying equilibria
Once we have equilibria, we can classify them using `phaseR::stability`.  There's a good deal of information beyond the `$classification` that we call and with `summary = T`.
```{r}
for (i in 1:nrow(eqs)) {
  print(x = paste0("x = ", eqs[i, 1], ", y = ", eqs[i, 2], " is a ", stability(deriv = model.ex1, ystar = eqs[i, ], parameters = model.parms, summary = F)$classification))
}
```

This will reveal our three interior equilibria---two stable and one unstable equilibrium.  Our current focus is on determining the relative stability around the two interior equilibria when stochasticity is added to the system.  In the next vignette, we heuristically run some stochastic simulations to better understand how the system behaves with stochasticity.
