---
title: "QPot: An R Package for Stochastic Differential Equation Quasi-Potential Analysis"
author:
  - "Christopher Moore, Chris Stieha, Ben Nolting, Maria Cameron, and Karen Abbott"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{2. Stochastic simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
model.ex1 <- function(t, y, parameters) {
  x <- y[1]
  y <- y[2]
  alpha <- parameters["alpha"]
  beta <- parameters["beta"]
  delta <- parameters["delta"]
  kappa <- parameters["kappa"]
  gamma <- parameters["gamma"]
  mu <- parameters["mu"]
  dy <- numeric(2)
  dy[1] <- (alpha * x) * (1 - (x / beta)) - ((delta * (x^2) * y) / (kappa + (x^2)))
  dy[2] <- ((gamma * (x^2) * y) / (kappa + (x^2))) - mu * (y^2)
  list(dy)
}
model.parms <- c("alpha" = 1.54, "beta" = 10.14, "delta" = 1, "gamma" = 0.476, "mu" = 0.112509, "kappa" = 1)
library(package = "rootSolve")
xspace <- seq(from = 1, to = 5, length.out = 10)
yspace <- seq(from = 2.5, to = 4, length.out = 10)
l.xspace <- length(x = xspace)
l.yspace <- length(x = yspace)
space.mat <- matrix(data = NA, nrow = l.xspace * l.yspace, ncol = 2)

for (i in 1:l.xspace) {
  for (j in 1:l.yspace) {
    y <- c(x = xspace[i], y = yspace[j])
    STO <- stode(y = y, func = model.ex1, parms = model.parms, positive = T)
    space.mat[(((i - 1) * l.xspace) + j), ] <- STO$y
  }
}
eqs <- unique(x = round(x = space.mat, digits = 3))
```


Stochastic simulation
========================================================
## Example 1 from Moore et al. (2016)
### Creating a model with stochasticity
In the previous vignette we used an example of a coupled consumer-resource equation developed for plankton and their consumers.  We now write the equation as a stochastic differential equation:

$$\mathrm{d}X(t) = \left(\alpha X(t)\left(1 - \frac{X(t)}{\beta}\right) - \frac{\delta X^2(t)Y(t)}{\kappa + X^2(t)}\right)dt + \sigma \mathrm{d}W_1(t)$$
$$\mathrm{d}Y(t) = \left(\frac{\gamma X^2(t)Y(t)}{\kappa + X^2(t)} - \mu Y^2(t)\right) + \sigma \mathrm{d}W_2(t)$$

There are several existing packages in R that will run stochastic simulations (see references in [CRAN Task View: Differential Equations](https://cran.r-project.org/web/views/DifferentialEquations.html)), but we include a less-computationally-efficient function for less-intensive simulations that are sufficient for our purposes and takes arguments in the form that we later use for the quasi-potential analysis.

### Running the simulation
Specifically, we created `TSTraj` (for **T**ime **S**eries **Traj**ectory), to take a equations as strings, with the option to specify the parameter values as a separate list.  `TSTraj` allows users to add stochasticity to the model (`sigma`), control  the time step ($\Delta t$), set upper and lower bounds (e.g., a lower bound of 0 may be useful for biologists studying populations since a population of < 0 is yet to be discovered), and returns a matrix with the numbers of rows equal to the total time steps ($\Delta t \times T$) and two columns for each state variable (e.g., $X$ and $Y$).

Specifying the above equations can be separate from the parameters:
```{r}
library(QPot)
var.eqn.x <- "(alpha * x) * (1 - (x / beta)) - ((delta * (x^2) * y) / (kappa + (x^2)))"
var.eqn.y <- "((gamma * (x^2) * y) / (kappa + (x^2))) - mu * (y^2)"
model.parms <- c(alpha = 1.54, beta = 10.14, delta = 1, gamma = 0.476, kappa = 1, mu = 0.112509)
```
or together, using `QPot::Model2String`
```{r}
parms.eqn.x <- Model2String(model = var.eqn.x, parms = model.parms, supress.print = T)
parms.eqn.y <- Model2String(model = var.eqn.y, parms = model.parms, supress.print = T)
```
For this simulation, we start our simulation at (1, 2), add Gaussian noise with a mean of 0 and a standard deviation of 0.05 every $\Delta t = 0.025$, for a total time of $T = 1000$.
```{r}
model.state <- c(x = 1, y = 2)
model.sigma <- 0.05
model.time <- 1000
model.deltat <- 0.025

set.seed(6174)
ts.ex1 <- TSTraj(y0 = model.state, time = model.time, deltat = model.deltat, x.rhs = var.eqn.x, y.rhs = var.eqn.y, sigma = model.sigma, parms = model.parms)
```
Or alternatively, one could also use TSTraj to combine equation strings and parameter values.
```{r}
ts.ex1 <- TSTraj(y0 = model.state, time = model.time, deltat = model.deltat, x.rhs = parms.eqn.x, y.rhs = parms.eqn.y, parms = model.parms, sigma = model.sigma)
```

### Visualizing the results
We have two functions that allow users to see the results from `TSTraj`: `TSPlot` and `TSDensity`.  First, `TSPlot` plots the time series for each state variable if `dim = 1` (default) and shows the trajectory in state space if `dim = 2`.  For `dim = 1`, we provide an option (default) to plot the density of each state variable adjacent to the time series plot.
```{r}
TSPlot(mat = ts.ex1, deltat = model.deltat)
TSPlot(mat = ts.ex1, deltat = model.deltat, dim = 2)
```
Second, `TSDensity` takes the simulation results and plots it as either a single (`dim = 1`) dimension or in two dimensional state space (`dim = 2`).
```{r}
TSDensity(mat = ts.ex1, dim = 1)
TSDensity(mat = ts.ex1, dim = 2)
```

### Discrepancies in the notion of stability in stochastic systems
We can see from these simulations that the system spends a great deal of time around the stable focus at x = 1.405 and y = 2.808 than the stable node at x = 4.904 and y = 4.062.  This realization is typical of this system, regardless of $\Delta t$, $T$, or $\sigma$.  We should therefore describe the behavior of system in a way that captures this behavior.  

The typical way of describing stability is through linear stability analysis, which respectively yields the following results for stable focus and stable node:
```{r}
library(phaseR)
stability(deriv = model.ex1, ystar = eqs[1, ], parameters = model.parms, summary = F)$eigenvalues
stability(deriv = model.ex1, ystar = eqs[3, ], parameters = model.parms, summary = F)$eigenvalues
```
The largest real eigenvalue of the Jacobian matrix for the stable focus is -0.0473848 and for the stable node is -0.37737660.  Because the stable node is larger (i.e., more negative), **we would conclude that the stable node is more stable than the stable focus---in direct contrast to what our simulation shows**.  This type of stability---known as asymptotic stability---is not necessarily sufficient for describing stability in stochastic systems, with continual perturbations.

This is why we need another tool that will better describe the behavior of how a system will behave with continual perturbations.  For this, Nolting and Abbott (2016) have argued that the quasi-potential should be used.

For a more in-depth description and mathematical details, we again encourage readers to see

> B. C. Nolting and K. C. Abbott. Balls, cups, and quasi-potentials: Quantifying stability in stochastic systems. Ecology, 97(4):850–864, 2016.

and specifically **$\S$ A Path Through the Quagmire of Stability Concepts**.  
